### 2023-06-21 submission/manuscript.tex
Ready to submit

Over the past 10 years, the best method of normalizing the sequencing
depth of samples characterized by 16S rRNA gene sequencing has been
contentious. An often cited paper by McMurdie and Holmes forcefully
argued that rarefying the number of sequence counts was ``inadmissible''
and should not be employed. However, I identified a number of problems
with the design of their simulations and analysis that could compromise
their results. In fact, when I reproduced and expanded upon their
analysis, it was clear that rarefaction was actually the most robust
approach for controlling for uneven sequencing effort across samples.
Rarefaction limits the rate of falsely detecting and rejecting
differences between treatment groups. Far from being ``inadmissible'',
rarefaction is a valuable tool for analyzing microbiome sequence data.

@@

In 2014 Paul McMurdie and Susan Holmes published ``Waste not, want not:
why rarefying microbiome data is inadmissible'' (WNWN) in \emph{PLOS
Computational Biology} (24). Their paper has had a significant impact on
the approaches that microbiome researchers use to analyze 16S rRNA gene
sequence data. According to Google Scholar, this paper has been cited
more than 2,400 times as of May 2023. There has been a rebuttal of WNWN
that showed how rarefaction is beneficial in some cases (25). I have
received correspondence from researchers over the past 10 years asking
how to address critiques from reviewers who criticize my correspondents'
analysis for using rarefaction (e.g., see
\url{https://twitter.com/inanna_nalytica/status/1264299305067786242}). I
have also received similar comments from reviewers regarding my own
work. For example, I received the critique for a preprint that I posted
in 2020 that analyzed the practice of removing rare taxa from microbiome
analyses {[}see comments linked to (4);
\url{https://web.archive.org/web/20230308191505/http://dalmug.org/filter-review/}{]}.
In the process of responding to these reviewers' comments and preparing
a manuscript investigating rarefaction and other approaches to control
for uneven sequencing effort, I grew to appreciate that there is
significant confusion in the field over what is meant by ``rarefying''
and ``rarefaction''.

@@

McMurdie and Holmes analyzed the effect of subsampling and other
approaches on clustering accuracy using what they called ``Simulation
A'' in their Figure 2A and elsewhere in WNWN. In Simulation A, they
investigated the ability to correctly assign simulated microbiome
samples to one of two clusters representing two simulated treatment
groups. Thankfully, the R code for Simulation A was provided by the
authors in the
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
R-flavored markdown file that was published as Protocol S1 in WNWN. I
will outline the simulation strategy below and will reference line
numbers from their R-flavored markdown file with ``L'' as a prefix.

@@

Non-normalized data were used to calculate Bray-Curtis, Euclidean,
Poisson, and Weighted UniFrac distances (L793-798, L801-806). Relative
abundance data were used to calculate Bray-Curtis, Unweighted UniFrac ,
and Weighted UniFrac distances (L760-767). Subsampled data were used as
input to calculate distances between samples using Bray-Curtis,
Euclidean, Unweighted UniFrac, Weighted UniFrac distances, Poisson
distances, and top-Mean Squared Difference (L769-775, L793-798,
L801-806). DESeq Variance Stabilization normalized data were used to
calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances
(L793-798). The Upper-Quartile Log-Fold Change normalized data were only
used to calculate top-mean squared difference distances (L801-806). WNWN
calculated Bray-Curtis (31), Euclidean (31), Unweighted UniFrac (32),
and Weighted UniFrac distances (33) using the \texttt{Phyloseq} R
package (34). They calculated Poisson distances using the
\texttt{PoiClaClu} R package (35). The top-mean squared difference was
calculated using the \texttt{edgeR} R package (19). The resulting
distance matrices were used to cluster the 80 samples into one of two
clusters using partitioning around the medoid (PAM), K-means clustering,
and hierarchical clustering (L865-879). Although data for all three
methods were presented in Protocol S1, only the PAM data were presented
in the main manuscript. The accuracy of the clustering assignments was
quantified as the fraction of the 80 samples that were assigned to the
correct cluster (L887-908). Since some of the subsampling conditions
removed samples below a minimum sequencing depth threshold, those
samples were counted as mis-clustered samples yielding minimum
accuracies below 50\%.

@@

Before assessing the impact of the points I critiqued above, I attempted
to replicate the results shown in Figures 4 and 5 of WNWN using the
authors' code. I created a Conda environment that used the R version and
package versions that were as close as possible to those used in WNWN.
It was necessary to patch the WNWN's R-flavored markdown file to render
the document to be compatible with the overall workflow of this study. I
was able to generate figures similar to those presented as WNWN's
Figures 4 and 5; my results are shown in this study as Figures 1 and 2,
respectively. The differences in results are likely due to differences
in software versions and operating systems. It is also worth noting that
the published versions of the two figures differ from those included in
Protocol S1 within the rendered HTML file
(\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.html})
and that the figure numbers are one higher in the paper than those
generated by the R-flavored markdown file (i.e., Protocol S1 labels the
figures as Figures 3 and 4 corresponding to the published Figures 4 and
5). Regardless of the differences, my results were qualitatively similar
to that provided in WNWN's Protocol S1.

@@

Each simulated condition was replicated 5 times in WNWN and the paper
reported the mean and standard deviation of the replicate clustering
accuracies. The relatively small number of replicates accounts for the
jerkiness of the lines in WNWN's Figures 4 and 5 (e.g.~the Bray-Curtis
distances calculated on the DESeqVS normalized data). A better approach
would have been to use 100 replicates as this would reduce the
dependency of the results on the random number generator's seed. By
increasing the number of replicates it was also possible to compare the
probability of falsely and correctly clustering samples from the same
and different treatment groups together (see points 10 and 11, below).
Because the accuracies were not symmetric around the mean accuracy
values the median and 95\% confidence intervals or interquartile range
should have been reported. To test the effect of increasing the number
of replicates, I pulled apart the code in
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
into individual R and bash scripts that were executed using a Snakemake
workflow with the same Conda environment I used above. This was
necessary since the number of simulated conditions increased 20-fold
with the additional replicates. Such intense data processing was not
practical within a single R-flavored markdown document. Again, the
observed results were qualitatively similar to those generated using the
single R-flavored markdown file (Figures 3 and 4). The increased number
of replications resulted in smoother lines and allowed me to present
empirical 95\% confidence intervals. For all analyses in the remainder
of this paper, I used 100 randomized replicates per condition.

@@

Close comparison of WNWN's Figure 4 and my version (Figure 3) revealed
several difference between the two plots. First, in the WNWN analysis,
the accuracies for the Weighted UniFrac distances at the largest effect
size (i.e., 3.5) were 1.00 for median sequencing depths of 1,000, 2,000,
and 10,000. In my version of the analysis, the values for the same
sequencing depths were 1.00, 1.00, and 1.00, respectively. The 95\%
confidence interval for these accuracies spanned between 0.51 and 1.00.
Second, clustering accuracy for Bray-Curtis and Weighted UniFrac
distances calculated using DESeq Variance Stabilization normalized OTU
counts were different between the original and my simulation at smaller
effect sizes and had wide confidence intervals. Inspection of the DESeq
Variance Stabilization normalized OTU counts revealed that the method
resulted in negative values. It has been suggested that WNWW turned
negative DESeq normalized counts to zero (25); however, I was unable to
find code in \texttt{simulation-cluster-accuracy-server.Rmd} that made
this transformation. In fact, rendering the R-flavored markdown files in
WNWN's Protocol S1 generated warning messages when passing the DESeq
normalized counts to the Bray-Curtis calculator, which said, ``results
may be meaningless because data have negative entries in method
`bray'\,''. Although the Weighted UniFrac calculator function did not
generate a similar warning message, negative count values would also
result in similarly meaningless distances. Both are due to the fact that
the distance calculators sum the counts of each OTU in both samples
being compared. In contrast, a Euclidean distance does not use a similar
sum, but sums the square of the difference between the OTU abundance in
each sample. Even if negative values were converted to zeroes, this
would effectively the same as removing rare taxa, which could have a
significant impact on the shape of the communities (4). To assess the
prevalence of negative counts in the simulated data, I quantified the
fraction of negative values in the OTU matrix from each simulation and
counted the number of simulations where the normalized OTU table had at
least one negative value (Figure 5). In general the fraction of negative
OTU counts increased with effect size, but decreased with sequencing
effort. The fraction of simulations with at least one negative value
increased with effect size and sequencing effort. The high frequency of
negative OTU counts resulted in highly variable Bray-Curtis and Weighted
UniFrac values. It is likely that because the WNWN analysis only used 5
replicates that the large variation in accuracies at high effect sizes
was missed initially. For the rest of this reanalysis study, I will only
report results using the DESeq-based Variance Stabilization
normalization with the Euclidean distance.

@@

The clustering accuracy measurements reported in the body of WNWN were
determined using PAM-based clusters while Protocol S1 also includes
K-means and hierarchical clustering. Although the data were not
displayed in a manner that lent itself to direct comparison in Protocol
S1, close inspection of the rendered figures suggested that PAM may not
have been the optimal choice in all situations. Rather, K-means
clustering appeared to perform better in many simulations. Because the
accuracies were the smallest at lower effect sizes, I focused my
comparison at the effect size of 1.15. For each set of 100 replicated
simulated datasets, I compared the clustering accuracy across clustering
methods to see how often each clustering method resulted in the highest
accuracy (Figure 7). Indeed, K-means clustering performed better than
the other methods. Among all combinations of normalization methods,
distance calculations, and read depths, K-means clustering was at least
as good as the other methods in 74.39\% of the randomizations (Figure
7); PAM clustering resulted in clustering accuracies as good or better
than the other methods in 49.92\% of the randomizations; and
hierarchical clustering (HClust) was at least as good as the other
methods for 44.32\% of the randomizations. Finally, I specifically
compared the clustering accuracies using rarefaction for each of the
distance calculations methods using PAM and K-means clustering. Among
the 30 combinations of distance calculations and read depths, K-means
performed better than PAM in 29 cases with PAM doing better in the 1
other case (i.e., calculating distances with Euclidean using 10,000
sequences). When using subsampled data, K-means clustering performed
better than PAM in each case. Because K-means clustering did so much
better than PAM clustering in the simulated conditions, I used K-means
clustering for the remainder of this study.

@@

In WNWN, the authors quantified the tradeoff between median sequencing
depth, the number of samples removed below the threshold, and clustering
accuracy (WNWN's Figure 5, my Figure 4). Although the optimal threshold
varied by distance metric, normalization method, and sequencing depth,
they removed samples whose number of sequences was less than the 15th
percentile (L404-419). They acknowledged that this screening step, which
was only used with subsampling, would decrease clustering accuracy
putting it at a relative disadvantage to the other methods (page 5,
column 1, last paragraph). Therefore, it was not surprising that the
peak clustering accuracy for their subsampled data was at 85\%. Because
the true best result would not known \emph{a priori} in an actual
microbiome study, it would be impossible for researchers to conduct a
sensitivity analysis comparing the tradeoffs between sequencing depth,
sample number, and clustering accuracy to select a sequencing depth for
their analysis without the risk of P-hacking. The differences in
clustering accuracy between subsampling and rarefaction with PAM and
K-means clustering indicated that it was necessary to reassess the
tradeoff between the library size minimum quantile and clustering
accuracy. When using rarefaction, K-means clustering, and only
considering conditions with 2,000 or more sequences, there was not a
condition where setting a higher threshold resulted in a better accuracy
than using all of the samples (Figure 8). These results showed that for
modern sequencing depths, using the full datasets with rarefaction and
K-means clustering resulted in accuracies that were better than those
observed when removing the smallest 15\% of the samples from each
simulated dataset. When the WNWN Figure 4 was recast with these
approaches, rarefaction performed at least as well as any of the other
transformations with each distance calculation (Figure 9).

@@

I and others have observed that not using rarefaction can lead to
falsely detecting differences between communities when sequencing effort
is confounded with the treatment group (25, 40). Previous analyses
showed that in these situations rarefaction did the best job of
controlling the rates of false detection (i.e., Type I errors) and
maintaining the statistical power to detect differences (i.e., 1-rate of
Type II errors) of differences between groups of samples. Such
situations have been observed when comparing communities at different
body parts where one site is more likely to generate contaminating
sequence reads from the host (41). To determine whether this result was
replicated with the WNWN simulation framework, I created a sequencing
depth distribution where sequencing depth was fully confounded with the
treatment group using both the GlobalPatterns and Log-scaled sequence
distributions. To confound the sequencing depth, sequencing depths from
one treatment group were drawn from below the median number of sequences
of the sequencing distribution and those for the second treatment group
were from above the median. To assess the risk of falsely detecting
clusters, I compared the clustering accuracies using an effect size of
1.00 using both the confounded and randomized sequencing distributions
(see rows 1 and 2 of Figure 14). The samples should have only been
assigned to one cluster; however, each of the clustering methods forced
the samples into two clusters. So, when there are two groups of 40
samples that do not differ, the best a method could do would be to
correctly assign 41 of the 80 samples for an accuracy of 0.51. The false
detection risk did not vary by method when the sequencing depth was
randomized across treatment groups. Yet, when the sequencing depth was
confounded with treatment group, rarefaction was the most consistent
normalization method for controlling the generation of spurious
clusters. At larger effect sizes the ability to correctly identify two
clusters increased when the sequencing depth was confounded with
treatment group (Figure 14). At the effect size of 1.15, the rarefied
data generated the highest accuracy clusters regardless of whether the
sequencing depths were confounded. Although the level of confounding in
this simulation was extreme, it highlights the ability of rarefaction to
control the false detection rate and the ability to correctly detect
clusters relative to the other normalization methods.

@@

The conclusions from WNWN have had a lasting impact on how researchers
analyze microbiome sequence data. As I have demonstrated using WNWN's
simulation framework, their claims are not supported. The most important
points that lead to the difference in our conclusions include the choice
of clustering algorithm and arbitrarily selecting a sequencing effort
threshold that was used to remove samples. Furthermore, the decision to
evaluate the effectiveness of normalization methods based on clustering
samples adds a layer of analysis that has been controversial and not
widely used. Ultimately, the authors choice of the word ``rarefying''
has sewn confusion in the field because it is often used in place of the
word ``rarefaction''. As I have demonstrated there were numerous choices
throughout the original study that resulted in rarefying/subsampling
looking worse than it truly is. In fact, when the data of the current
study are taken as a whole, rarefaction is the preferred approach. Short
of obtaining the exact same number of sequences from each sample,
rarefaction remains the best approach to control for uneven sequencing
effort when analyzing alpha and beta diversity metrics. Although beyond
the scope of the current study, the same principles would support using
rarefaction to control for uneven sequencing effort in shotgun
metagenomic sequencing analyses.

Beyond the discussion of whether rarefaction is appropriate for
analyzing microbiome data it is worth commenting on WNWN's advice to use
DESeq's Variance Stabilization or edgeR's Upper-Quartile Log-fold Change
normalization strategies. These methods have been adopted from gene
expression analysis to microbiome analysis. Gene expression analysis
implicitly assumes that all samples have the same genes since. While
this might work in comparing healthy and diseased tissues from a cohort
of patients, it does not generalize to those patients' microbiota.
Microbial populations are highly patchy in their distribution. Thus, a
zero count for gene expression is more likely to represent a gene below
the limit of detection whereas a zero count for a microbiome analysis is
more likely to represent the true absence of the OTU. In fact, zeroes
are less common in gene expression analyses. But because microbiome
studies have so many zeroes, it is necessary to add a pseudo-count to
all OTUs for both the edgeR and DESeq-based normalizations (L443 and
L487, respectively). In WNWN, a pseudo-count of 1 was used. However,
this value is arbitrary and the results can vary based on the choice of
pseudo-count and the patchiness of the communities being analyzed. Since
WNWN was published, compositional approaches have been proposed to
account for uneven sequencing and to provide improved interpretability
(14, 16--18, 46, 47). However, these methods also often require the use
of pseudo-counts and are not actually insensitive to uneven sequencing
(40). Rarefaction is preferred to these alternatives.

The choice of distance metric is a complicated question and the use of
six different metrics in WNWN illustrates the challenges of selecting
one. Within the ecology literature, Euclidean distances are widely
avoided because joint absence is weighted the same as joint presence of
taxa (31). As discussed in point 11, metrics that are based on community
membership (i.e., the presence or absence of taxa) performed worse than
those than those that were based on community structure (i.e., their
relative abundance). For this reason, the Unweighted UniFrac and other
metrics like the Jaccard or Sorenson distance coefficients should likely
be avoided. The Poisson distance metric is largely novel to the
microbial ecology literature and performed no better than the more
traditional metrics. In the current analysis, the phylogenetic Weighted
UniFrac distance performed comparably to Bray-Cutis distances for
clustering or differentiating between communities with adonis.
Regardless of the distance calculation employed, they all performed best
when using rarefaction relative to the other normalization methods.

Although the authors claim that ``Rarefying counts requires an arbitrary
selection of a library size minimum that affects downstream inference''
(page 8, column 1, point 3), in actual microbiome studies the selection
of a sequencing depth is not as arbitrary as the authors claim. Rather,
to avoid ``P-hacking'', researchers pick a set of criteria where they
will include or exclude samples prior to testing their data. Examples of
criteria might include the presence of a large gap in the sequencing
effort distribution, a desire to include poorly sequenced controls, or
the \emph{a priori} stipulation of a minimum sequencing effort. To
mitigate concerns of arbitrary or engineered minimum library sizes,
researchers should indicate the rationale for the threshold they
selected.

>>>>

Over the past 10 years, the best method for normalizing the sequencing
depth of samples characterized by 16S rRNA gene sequencing has been
contentious. An often cited paper by McMurdie and Holmes forcefully
argued that rarefying the number of sequence counts was ``inadmissible''
and should not be employed. However, I identified a number of problems
with the design of their simulations and analysis that could compromise
their results. In fact, when I reproduced and expanded upon their
analysis, it was clear that rarefaction was actually the most robust
approach for controlling for uneven sequencing effort across samples.
Rarefaction limits the rate of falsely detecting and rejecting
differences between treatment groups. Far from being ``inadmissible'',
rarefaction is a valuable tool for analyzing microbiome sequence data.

@@

In 2014 Paul McMurdie and Susan Holmes published ``Waste not, want not:
why rarefying microbiome data is inadmissible'' (WNWN) in \emph{PLOS
Computational Biology} (24). Their paper has had a significant impact on
the approaches that microbiome researchers use to analyze 16S rRNA gene
sequence data. According to Google Scholar, this paper has been cited
more than 2,400 times as of June 2023. There has been a rebuttal of WNWN
that showed how rarefaction is beneficial in some cases (25); however,
the proponents of WNWN appear to be holding sway over the microbiome
community. I have received correspondence from researchers over the past
10 years asking how to address critiques from reviewers who criticize my
correspondents' analysis for using rarefaction (e.g., see
\url{https://twitter.com/inanna_nalytica/status/1264299305067786242}). I
have also received similar comments from reviewers regarding my own
work. For example, I received the critique for a preprint that I posted
in 2020 that analyzed the practice of removing rare taxa from microbiome
analyses {[}see comments linked to (4);
\url{https://web.archive.org/web/20230308191505/http://dalmug.org/filter-review/}{]}.
In the process of responding to these reviewers' comments and preparing
a manuscript investigating rarefaction and other approaches to control
for uneven sequencing effort, I grew to appreciate that there is
significant confusion in the field over what is meant by ``rarefying''
and ``rarefaction''.

@@

McMurdie and Holmes analyzed the effect of subsampling and other
approaches on clustering accuracy using what they called ``Simulation
A'' in their Figure 2A and elsewhere in WNWN. In Simulation A, they
investigated the ability to correctly assign simulated microbiome
samples to one of two clusters representing two simulated treatment
groups. Thankfully, the R code for Simulation A was provided by the
authors in the
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
R-flavored markdown (R markdown) file that was published as Protocol S1
in WNWN. I will outline the simulation strategy below and will reference
line numbers from their R markdown file with ``L'' as a prefix.

@@

Non-normalized data were used to calculate Bray-Curtis, Euclidean,
Poisson, and Weighted UniFrac distances (L793-798, L801-806). Relative
abundance data were used to calculate Bray-Curtis, Unweighted UniFrac ,
and Weighted UniFrac distances (L760-767). Subsampled data were used as
input to calculate distances between samples using Bray-Curtis,
Euclidean, Unweighted UniFrac, Weighted UniFrac distances, Poisson
distances, and top-Mean Squared Difference (L769-775, L793-798,
L801-806). DESeq Variance Stabilization normalized data were used to
calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances
(L793-798). The Upper-Quartile Log-Fold Change normalized data were only
used to calculate top-Mean Squared Difference distances (L801-806). WNWN
calculated Bray-Curtis (31), Euclidean (31), Unweighted UniFrac (32),
and Weighted UniFrac distances (33) using the \texttt{Phyloseq} R
package (34). They calculated Poisson distances using the
\texttt{PoiClaClu} R package (35). The top-mean squared difference was
calculated using the \texttt{edgeR} R package (19). The resulting
distance matrices were used to cluster the 80 samples into one of two
clusters using partitioning around the medoid (PAM), K-means clustering,
and hierarchical clustering (L865-879). Although data for all three
methods were presented in Protocol S1, only the PAM data were presented
in the main manuscript. The accuracy of the clustering assignments was
quantified as the fraction of the 80 samples that were assigned to the
correct cluster (L887-908). Since some of the subsampling conditions
removed samples below a minimum sequencing depth threshold, those
samples were counted as mis-clustered samples yielding minimum
accuracies below 50\%.

@@

Before assessing the impact of the points I critiqued above, I attempted
to replicate the results shown in Figures 4 and 5 of WNWN using the
authors' code. I created a Conda environment that used the R version and
package versions that were as close as possible to those used in WNWN.
It was necessary to patch the WNWN's R markdown file to render the
document to be compatible with the overall workflow of this study. I was
able to generate figures similar to those presented as WNWN's Figures 4
and 5; my results are shown in this study as Figures 1 and 2,
respectively. The differences in results are likely due to differences
in software versions and operating systems. It is also worth noting that
the published versions of the two figures differ from those included in
Protocol S1 within the rendered HTML file
(\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.html})
and that the figure numbers are one higher in the paper than those
generated by the R markdown file (i.e., Protocol S1 labels the figures
as Figures 3 and 4 corresponding to the published Figures 4 and 5).
Regardless of the differences, my results were qualitatively similar to
that provided in WNWN's Protocol S1.

@@

Each simulated condition was replicated 5 times in WNWN and the paper
reported the mean and standard deviation of the replicate clustering
accuracies. The relatively small number of replicates accounts for the
jerkiness of the lines in WNWN's Figures 4 and 5 (e.g.~the Bray-Curtis
distances calculated on the DESeqVS normalized data). A better approach
would have been to use 100 replicates as this would reduce the
dependency of the results on the random number generator's seed. By
increasing the number of replicates it was also possible to compare the
probability of falsely and correctly clustering samples from the same
and different treatment groups together (see points 10 and 11, below).
Because the accuracies were not symmetric around the mean accuracy
values the median and 95\% confidence intervals or interquartile range
should have been reported. To test the effect of increasing the number
of replicates, I pulled apart the code in
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
into individual R and bash scripts that were executed using a Snakemake
workflow with the same Conda environment I used above. This was
necessary since the number of simulated conditions increased 20-fold
with the additional replicates. Such intense data processing was not
practical within a single R markdown document. Again, the observed
results were qualitatively similar to those generated using the single R
markdown file (Figures 3 and 4). The increased number of replications
resulted in smoother lines and allowed me to present empirical 95\%
confidence intervals. For all analyses in the remainder of this paper, I
used 100 randomized replicates per condition.

@@

Close comparison of WNWN's Figure 4 and my version (Figure 3) revealed
several difference between the two plots. First, in the WNWN analysis,
the accuracies for the Weighted UniFrac distances at the largest effect
size (i.e., 3.5) were 1.00 for median sequencing depths of 1,000, 2,000,
and 10,000. In my version of the analysis, the values for the same
sequencing depths were 1.00, 1.00, and 1.00, respectively. The 95\%
confidence interval for these accuracies spanned between 0.51 and 1.00.
Second, clustering accuracy for Bray-Curtis and Weighted UniFrac
distances calculated using DESeq Variance Stabilization normalized OTU
counts were different between the original and my simulation at smaller
effect sizes and had wide confidence intervals. Inspection of the DESeq
Variance Stabilization normalized OTU counts revealed that the method
resulted in negative values. It has been suggested that WNWW turned
negative DESeq normalized counts to zero (25); however, I was unable to
find code in \texttt{simulation-cluster-accuracy-server.Rmd} that made
this transformation. In fact, rendering the R markdown files in WNWN's
Protocol S1 generated warning messages when passing the DESeq normalized
counts to the Bray-Curtis calculator, which said, ``results may be
meaningless because data have negative entries in method `bray'\,''.
Although the Weighted UniFrac calculator function did not generate a
similar warning message, negative count values would also result in
similarly meaningless distances. Both are due to the fact that the
distance calculators sum the counts of each OTU in both samples being
compared. In contrast, a Euclidean distance does not use a similar sum,
but sums the square of the difference between the OTU abundance in each
sample. Even if negative values were converted to zeroes, this would
effectively the same as removing rare taxa, which could have a
significant impact on the shape of the communities (4). To assess the
prevalence of negative counts in the simulated data, I quantified the
fraction of negative values in the OTU matrix from each simulation and
counted the number of simulations where the normalized OTU table had at
least one negative value (Figure 5). In general the fraction of negative
OTU counts increased with effect size, but decreased with sequencing
effort. The fraction of simulations with at least one negative value
increased with effect size and sequencing effort. The high frequency of
negative OTU counts resulted in highly variable Bray-Curtis and Weighted
UniFrac values. It is likely that because the WNWN analysis only used 5
replicates that the large variation in accuracies at high effect sizes
was missed initially. For the rest of this reanalysis study, I will only
report results using the DESeq-based Variance Stabilization
normalization with the Euclidean distance.

@@

The clustering accuracy measurements reported in the body of WNWN were
determined using PAM-based clusters while Protocol S1 also includes
K-means and hierarchical clustering. Although the data were not
displayed in a manner that lent itself to direct comparison in Protocol
S1, close inspection of the rendered figures suggested that PAM may not
have been the optimal choice in all situations. Actually, K-means
clustering appeared to perform better in many simulations. Because the
accuracies were the smallest at lower effect sizes, I focused my
comparison at the effect size of 1.15. For each set of 100 replicated
simulated datasets, I compared the clustering accuracy across clustering
methods to see how often each clustering method resulted in the highest
accuracy (Figure 7). Indeed, K-means clustering performed better than
the other methods. Among all combinations of normalization methods,
distance calculations, and read depths, K-means clustering was at least
as good as the other methods in 74.39\% of the randomizations (Figure
7); PAM clustering resulted in clustering accuracies as good or better
than the other methods in 49.92\% of the randomizations; and
hierarchical clustering (HClust) was at least as good as the other
methods for 44.32\% of the randomizations. Finally, I specifically
compared the clustering accuracies using rarefaction for each of the
distance calculations methods using PAM and K-means clustering. Among
the 30 combinations of distance calculations and read depths, K-means
performed better than PAM in 29 cases with PAM doing better in the 1
other case (i.e., calculating distances with Euclidean using 10,000
sequences). When using subsampled data, K-means clustering performed
better than PAM in each case. Because K-means clustering did so much
better than PAM clustering in the simulated conditions, I used K-means
clustering for the remainder of this study.

@@

In WNWN, the authors quantified the tradeoff between median sequencing
depth, the number of samples removed below the threshold, and clustering
accuracy (WNWN's Figure 5, my Figure 4). Although the optimal threshold
varied by distance metric, normalization method, and sequencing depth,
they removed samples whose number of sequences was less than the 15th
percentile (L404-419). They acknowledged that this screening step, which
was only used with subsampling, would decrease clustering accuracy
putting it at a relative disadvantage to the other methods (page 5,
column 1, last paragraph). Therefore, it was not surprising that the
peak clustering accuracy for their subsampled data was at 85\%. Because
the true best threshold would not known \emph{a priori} in an actual
microbiome study, it would be impossible for researchers to conduct a
sensitivity analysis comparing the tradeoffs between sequencing depth,
sample number, and clustering accuracy to select a sequencing depth for
their analysis without the risk of p-hacking. The differences in
clustering accuracy between subsampling and rarefaction with PAM and
K-means clustering indicated that it was necessary to reassess the
tradeoff between the library size minimum quantile and clustering
accuracy. When using rarefaction, K-means clustering, and only
considering conditions with 2,000 or more sequences, there was not a
condition where setting a higher threshold resulted in a better accuracy
than using all of the samples (Figure 8). These results showed that for
modern sequencing depths, using the full datasets with rarefaction and
K-means clustering resulted in accuracies that were better than those
observed when removing the smallest 15\% of the samples from each
simulated dataset. When the WNWN Figure 4 was recast with these
approaches, rarefaction performed at least as well as any of the other
transformations with each distance calculation (Figure 9).

@@

I and others have observed that not using rarefaction can lead to
falsely detecting differences between communities when sequencing effort
is confounded with the treatment group (25, 40). Previous analyses
showed that in these situations rarefaction did the best job of
controlling the rates of false detection (i.e., Type I errors) and
maintaining the statistical power to detect differences (i.e., 1-rate of
Type II errors) of differences between groups of samples. Such
situations have been observed when comparing communities at different
body parts where one site is more likely to generate contaminating
sequence reads from the host (e.g., 41). To determine whether this
result was replicated with the WNWN simulation framework, I created a
sequencing depth distribution where sequencing depth was fully
confounded with the treatment group using both the GlobalPatterns and
Log-scaled sequence distributions. To confound the sequencing depth,
sequencing depths from one treatment group were drawn from below the
median number of sequences of the sequencing distribution and those for
the second treatment group were from above the median. To assess the
risk of falsely detecting clusters, I compared the clustering accuracies
using an effect size of 1.00 using both the confounded and randomized
sequencing distributions (see rows 1 and 2 of Figure 14). The samples
should have only been assigned to one cluster; however, each of the
clustering methods forced the samples into two clusters. So, when there
are two groups of 40 samples that do not differ, the best a method could
do would be to correctly assign 41 of the 80 samples for an accuracy of
0.51. The false detection risk did not vary by method when the
sequencing depth was randomized across treatment groups. Yet, when the
sequencing depth was confounded with treatment group, rarefaction was
the most consistent normalization method for controlling the generation
of spurious clusters. At larger effect sizes the ability to correctly
identify two clusters increased when the sequencing depth was confounded
with treatment group (Figure 14). At the effect size of 1.15, the
rarefied data generated the highest accuracy clusters regardless of
whether the sequencing depths were confounded. Although the level of
confounding in this simulation was extreme, it highlights the ability of
rarefaction to control the false detection rate and the ability to
correctly detect clusters relative to the other normalization methods.

@@

The conclusions from WNWN have had a lasting impact on how researchers
analyze microbiome sequence data. As I have demonstrated using WNWN's
simulation framework, their claims are not supported. The most important
points that lead to the difference in our conclusions include the choice
of clustering algorithm and arbitrarily selecting a sequencing effort
threshold that was used to remove samples. Furthermore, the decision to
evaluate the effectiveness of normalization methods based on clustering
samples adds a layer of analysis that has been controversial and not
widely used. Ultimately, the authors choice of the word ``rarefying''
has sewn confusion in the field because it is often used in place of the
word ``rarefaction''. As I have demonstrated there were numerous choices
throughout the original study that resulted in rarefying/subsampling
looking worse than it truly is. In fact, when the data of the current
study are taken as a whole, rarefaction is the preferred approach. Short
of obtaining the exact same number of sequences from each sample,
rarefaction remains the best approach to control for uneven sequencing
effort when analyzing alpha and beta diversity metrics. Although beyond
the scope of the current study, the same principles would also support
using rarefaction to control for uneven sequencing effort when
calculating alpha and beta diversity metrics in shotgun metagenomic
sequencing analyses.

Beyond the discussion of whether rarefaction is appropriate for
analyzing microbiome data it is worth commenting on WNWN's advice to use
DESeq's Variance Stabilization or edgeR's Upper-Quartile Log-fold Change
normalization strategies. These methods have been adopted from gene
expression analysis to microbiome analysis. Gene expression analysis
implicitly assumes that all samples have the same genes. While this
might work in comparing healthy and diseased tissues from a cohort of
patients, it does not generalize to those patients' microbiota.
Microbial populations are highly patchy in their distribution. Thus, a
zero count for gene expression is more likely to represent a gene below
the limit of detection whereas a zero count for a microbiome analysis is
more likely to represent the true absence of the OTU. In fact, zeroes
are less common in gene expression analyses. But because microbiome
studies have so many zeroes, it is necessary to add a pseudo-count to
all OTUs for both the edgeR and DESeq-based normalizations (L443 and
L487, respectively). In WNWN, a pseudo-count of 1 was used. However,
this value is arbitrary and the results can vary based on the choice of
pseudo-count and the patchiness of the communities being analyzed. Since
WNWN was published, compositional approaches have been proposed to
account for uneven sequencing and to provide improved interpretability
(14, 16--18, 46, 47). However, these methods also often require the use
of pseudo-counts and are not actually insensitive to uneven sequencing
(40). Rarefaction is preferred to these alternatives.

The choice of distance metric is a complicated question and the use of
six different metrics in WNWN illustrates the challenges of selecting
one. Within the ecology literature, Euclidean distances are widely
avoided because joint absence is weighted the same as joint presence of
taxa (31). As discussed in point 11, distances that are based on
community membership (i.e., the presence or absence of taxa) performed
worse than those than those that were based on community structure
(i.e., their relative abundance). For this reason, the Unweighted
UniFrac and other distances like the Jaccard or Sorenson distance
coefficients should likely be avoided. The Poisson distance metric is
largely novel to the microbial ecology literature and performed no
better than the more traditional distances. In the current analysis, the
phylogenetic Weighted UniFrac distance performed comparably to
Bray-Cutis distances for clustering or differentiating between
communities with \texttt{adonis2}. Regardless of the distance
calculation employed, they all performed best when using rarefaction
relative to the other normalization methods.

Although the authors claim that ``Rarefying counts requires an arbitrary
selection of a library size minimum that affects downstream inference''
(page 8, column 1, point 3), in actual microbiome studies the selection
of a sequencing depth is not as arbitrary as the authors claim. Rather,
to avoid p-hacking, researchers pick a set of criteria where they will
include or exclude samples prior to testing their data. Examples of
criteria might include the presence of a large gap in the sequencing
effort distribution, a desire to include poorly sequenced controls, or
the \emph{a priori} stipulation of a minimum sequencing effort. To
mitigate concerns of arbitrary or engineered minimum library sizes,
researchers should indicate the rationale for the threshold they
selected.

WNWN performed a second set of simulations to address the effect of
normalization method on the ability to correctly detect differential
abundance of OTUs that were randomly selected to have their relative
abundances changed (i.e., ``Simulation B''). Re-addressing this set of
simulations is beyond the scope of the current analysis and others have
already contributed critiques (25). However, many of the same concerns
addressed above would apply. Perhaps most important is the fact that if
the relative abundance of several OTUs increase, then the relative
abundance of all other OTUs would necessarily decrease because the data
are compositional (47). Thus, one would expect every OTU to be
differentially abundant. Because of this, it is not possible to truly
modify the abundance of a set of OTUs independent of all other OTUs.
This is an important limitation of tests of methods attempting to detect
differentially abundant OTUs. Regardless, a form of rarefaction could
still be employed for detecting differential abundance. One could
subsample the data, perform the statistical test, and identify the
differentially abundant OTUs. This process could be repeated and the
results synthesized. In my experience the largest differences between
subsamples are for low relative abundances OTUs, which are unlikely to
be statistically significant.
